Full Brain BOLD Signal Parameter Estimation \\ using Particle Filters

### 2010-08-28 doc/presentation/main.tex
more work getting the presentation togeth

\definecolor{mblue}{RGB}{178,34,34}
\definecolor{mred}{RGB}{0,0,128}
\definecolor{mgreen}{RGB}{0,100,0}

@@

\section{Introduction}
\begin{frame}{Balloon Model}
\begin{figure}
\includegraphics[width=10cm]{model}
\caption{
    \tiny
    \cite{Riera2003}
}
\end{figure}
\note{Rates from below 1 Hz to 5 to 6 Hundred Hz}
\note{Change in average firing rates increase metabolism and sympathetic response causes
        increased inflow of oxygenated blood.}
\note{In the version of the Balloon model I used, the flow is locked to metabolism.}
\note{According the balloon model this increases the local blood volume, and the outflow pressure.}
\note{So the signal is altered by changes in the ratio of Deoxyhemoglobin and Oxygenated Hemoglobin,
        which are a function of Volume and Oxygenation.}
\end{frame}

@@

%more expansion of all the parameters
\begin{frame}{BOLD State Space Model}
  \scriptsize
  \begin{columns}
  \begin{column}{.7\textwidth}
    \begin{itemize}
    \item State Vector:
    \begin{eqnarray}
    \color{mblue}\dot{s}(t) &=& {\color{mred}\epsilon} {\color{mgreen}u(t)} - 
                \frac{\color{mblue}s(t)}{\color{mred}\tau_s} - \frac{{\color{mblue}f(t)}-1}{\color{mred}\tau_f} \nonumber \\
    \color{mblue}\dot{f}(t) &=& \color{mblue}s \nonumber \\
    \color{mblue}\dot{v}(t) &=& \frac{{\color{mblue}f(t)} - 
                {\color{mblue}v(t)} ^ {1/{\color{mred}\alpha}}}{\color{mred}\tau_0} \nonumber \\
    \color{mblue}\dot{q}(t) &=& \frac{1}{\color{mred}\tau_0}
            \left(\frac{{\color{mblue}f(t)}(1-(1-{\color{mred}E_0})^{1/\color{mblue}f(t)})}{\color{mred}E_0} -
        \frac{\color{mblue}q(t)}{{\color{mblue}v(t)}^{1-1/\color{mred}\alpha}}\right) \nonumber 
    \end{eqnarray}
    \item Output:
    $$y(t) = {\color{mred}V_0}(a_1( 1 - {\color{mblue}Q(t)}) - a_2(1 - {\color{mblue}V(t)}))$$
    \end{itemize}
  \end{column}

@@

\begin{frame}{Nonlinear Methods}
\begin{itemize}
    \begin{columns}
    \begin{column}{.45\textwidth}
    \item Local Linearization filter, \cite{Riera2003}
    $$f(t)- f(t-1) \sim N(0, \sigma^2)$$
     \note{Jacobian of the $\dot{x}$ is available, making gradient descent possible}
    \item Genetic Algorithms and Simulated Annealing, \cite{Vakorin2007}
     \note{Genetic Algorithms were used to optimize spike points, simulated annealing for the parameters}
    \item Particle Filters to estimate $P(X_t | \Theta, Y_T)$ \cite{Murray2009}
    \begin{itemize}
        \item ML estimate of $\Theta$ based on $P(X_t | \Theta, Y_t)$, \cite{Johnston2007}
         \note{Particle Filter used to estimate states, ML of parameters based on the posterior of states and so on}
    \end{itemize}
    \item Unscented Kalman Filter to estimate parameters \cite{Hu2009}
     \note{Technically Kalman Filters and Particle Filters are Bayesian Filters}
     \note{Essentially filtering out inconsistent parameters.}
    \end{column}

    \begin{column}{.55\textwidth}
    \small
    \begin{figure}
    \includegraphics[width=.7\textwidth]{simulated_annealing}
    \caption{Simulated Annealing can escape local minima with chaotic jumps. \cite{Dama}}
    \end{figure}
    \begin{figure}
    \includegraphics[width=.7\textwidth]{kalman}
    \caption{(a) Initial Belief  (b) Noisy Measurement  (c) incorporated into the belief. \cite{Thrun2005}}
    \end{figure}
    \end{column}
    \end{columns}
\end{itemize}
\end{frame}

\subsection{Particle Filter}
%frame with example
\begin{frame}{Example System Identification}
\begin{itemize}
\item Given:
    \begin{itemize}
    \item Elevation Map (1D)
    \item Ability To Measure Elevation
    \end{itemize}
\item Particle Consists of:
    \begin{itemize}
    \item Location (Altered By Speed/Direction) $[0, 20]$
    \item Direction (Constant, Unknown) $\{Left, Right\}$
    \end{itemize}
\end{itemize}

@@

\begin{frame}{Construction}
\begin{itemize}
    \item Mixture PDF at time $k$, with $N_p$ particles:\\
        $$P(x_k) = \sum_{i=0}^{N_p} w^i\delta(x_k - x^i_k ) dx$$
    \item Particle Weights at time $t_k$:\\
        $$w^i_k \propto \frac{P(x^i_{0:k} | y_{0:k})}{q(x^i_{0:k} | y_{0:k})}$$
      \note{$q$ is the density/location of the particle}
      \note{$p$ is the actual probability }
    \item Incorporating Measurement:\\
        $$w^i_k \propto & w^i_{k-1}P(y_k| x_k) $$
      \note{Because of the proportion, note that the weights do not
            have to integrate to 1, and can be extremely large or extremely small}
\end{itemize}
\end{frame}

\begin{frame}{Particle Filter }
\footnotesize
\begin{itemize}
    \item 
\end{itemize}
\end{frame}

\section{Methods}
\begin{frame}{FMRI Noise}
\centering

@@

\section{Conclusion}
\begin{frame}{Factors Affecting Convergence}
    \begin{itemize}
        \item Weighting function
            \begin{itemize}
                \item Continuous, Long Tailed, Zero-Mean 
                \item Too wide $\rightarrow$ under-sensitivity, 
                    slow or no convergence
                \item Too thin $\rightarrow$ reduces robustness 
                    to noise, particle deprivation
            \end{itemize}
        \item Resampling 
            \begin{itemize}
                \item Stratified Resampling can result in truncated tails 
                        on posterior
                \item Regularized Resampling can result in over smoothing and 
                        slow convergence
            \end{itemize}
        \item Number of particles
            \begin{itemize}
                \item More particles give higher fidelity posterior
            \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{Conclusion}
\begin{itemize}
    \item Summary:
    \begin{itemize}
    \item BOLD Parameters Ill-Defined
    \item Particle Filter Capable of good parameter BOLD estimate with 1000 particles
    \item Mutual Information performs well as estimate of Quality
    \end{itemize}
    \item Future Works
    \begin{itemize}
        \item Further limitations should be placed on priors, Deneux 
            et al. \cite{Deneux2006} shows that parameters could imposed.
        \item Automation of Weighting Function
        \item Analysis of joint parameter distribution for populations
        \item Real Time analysis possible for multiple Voxels, similar to
            De Charms et al. \cite{DeCharms2005}

>>>>

\definecolor{mblue}{RGB}{178,34,34}
\definecolor{mred}{RGB}{0,0,128}
\definecolor{mgreen}{RGB}{0,100,0}
\definecolor{myellow}{RGB}{176,169,0}

@@

\section{Introduction}
\begin{frame}{Balloon Model}
\begin{figure}
\includegraphics[width=10cm]{model2}
\caption{
    \tiny
    \cite{Riera2003}
}
\end{figure}
\note{Its been known for about 20 years that changes in Blood Oxygen Level alter
        MR signals in certain types of imaging. The basis of contrast in fMRI is 
        changes in the so-called spin-spin relaxation (or T2$^*$ relaxation time
        of tissue). The rate of loss in nuclear spin coherence depends on the molecules in the voxel being
        measured \emph{and the surrounding voxels}. For this reason the signal 
        detected from BLOOD doesn't just depend on the amount of oxygen, but also the
        amount of De-oxygenated hemoglobin. The question then, is what drives changes
        in De-oxygenated Hemoglobin and blood volume?

        Well the brain has no local oxygen storage, so it is completely dependent on 
        fresh Oxygenated-Hemoglobin. When the brain becomes more active, firing rates
        for neurons increase by several orders of magnitude, and recharging these neurons
        requires additional oxygen for matabolism. }
\note{Rates from below 1 Hz to 5 to 6 Hundred Hz}
\note{Change in average firing rates increase metabolism and sympathetic response causes
        increased inflow of oxygenated blood.}
\note{In the version of the Balloon model I used, the flow is locked to metabolism.}
\note{According the balloon model this increases the local blood volume, and the outflow pressure.}
\note{So the signal is altered by changes in the ratio of Deoxyhemoglobin and Oxygenated Hemoglobin,
        which are a function of Volume and Oxygenation.}
\end{frame}

@@

%more expansion of all the parameters
\begin{frame}{BOLD State Space Model}
  \scriptsize
  \begin{columns}
  \begin{column}{.7\textwidth}
    \begin{itemize}
    \item State Vector:
    \begin{eqnarray}
    \color{mblue}\dot{s}(t) &=& {\color{mred}\epsilon} {\color{mgreen}u(t)} - 
                \frac{\color{mblue}s(t)}{\color{mred}\tau_s} - \frac{{\color{mblue}f(t)}-1}{\color{mred}\tau_f} \nonumber \\
    \color{mblue}\dot{f}(t) &=& \color{mblue}s \nonumber \\
    \color{mblue}\dot{v}(t) &=& \frac{{\color{mblue}f(t)} - 
                {\color{mblue}v(t)} ^ {1/{\color{mred}\alpha}}}{\color{mred}\tau_0} \nonumber \\
    \color{mblue}\dot{q}(t) &=& \frac{1}{\color{mred}\tau_0}
            \left(\frac{{\color{mblue}f(t)}(1-(1-{\color{mred}E_0})^{1/\color{mblue}f(t)})}{\color{mred}E_0} -
        \frac{\color{mblue}q(t)}{{\color{mblue}v(t)}^{1-1/\color{mred}\alpha}}\right) \nonumber 
    \end{eqnarray}
    \item Output:
    $${\color{myellow}y(t)} = {\color{mred}V_0}(a_1( 1 - {\color{mblue}Q(t)}) - a_2(1 - {\color{mblue}V(t)}))$$
    \end{itemize}
  \end{column}

@@

\begin{frame}{Nonlinear Methods}
\begin{itemize}
    \begin{columns}
    \begin{column}{.42\textwidth}
    \item Local Linearization filter, \cite{Riera2003}
    $$f(t)- f(t-1) \sim N(0, \sigma^2)$$
     \note{Jacobian of the $\dot{x}$ is available, making gradient descent possible}
    \item Genetic Algorithms and Simulated Annealing, \cite{Vakorin2007}
     \note{Genetic Algorithms were used to optimize spike points, simulated annealing for the parameters}
    \item Particle Filters to estimate States, \cite{Murray2009}
    \begin{itemize}
        \item ML estimate of $\Theta$, \cite{Johnston2007}
         \note{Particle Filter used to estimate states, ML of parameters based on the posterior of states and so on}
    \end{itemize}
    \item Unscented Kalman Filter \cite{Hu2009}
     \note{Technically Kalman Filters and Particle Filters are Bayesian Filters}
     \note{Essentially filtering out inconsistent parameters.}
    \end{column}

    \begin{column}{.58\textwidth}
    \begin{figure}
    \includegraphics[clip=true,trim=0cm 0cm 0cm 1cm,width=.6\textwidth]{simulated_annealing}
    \caption{Simulated Annealing can escape local minima with chaotic jumps. \cite{Dama}}
    \end{figure}
    \begin{figure}
    \includegraphics[width=\textwidth]{kalman}
    \caption{UKF: (a) Initial Belief  (b) Noisy Measurement  (c) incorporated into the belief. \cite{Thrun2005}}
    \end{figure}
    \end{column}
    \end{columns}
\end{itemize}
\end{frame}

\subsection{Particle Filter}
%frame with example
\begin{frame}{Example System Identification}
\begin{itemize}
\item Given:
    \begin{itemize}
    \item Elevation Map (1D)
    \item Ability To Measure Elevation
    \end{itemize}
\item Particle Consists of the unknowns:
    \begin{itemize}
    \item State: Location  $[0, 20]$
    \item Constant: Direction  $\{Left, Right\}$
    \end{itemize}
\end{itemize}

@@

\begin{frame}{Particle Filter Concepts}
\small
\begin{columns}
\begin{column}{.5\textwidth}
\begin{itemize}
    \item Weighting Function
    \item Resampling
    \begin{itemize}
        \item Remove Useless Particles
    \end{itemize}
    \item Regularized Resampling
    \begin{itemize}
        \item Prevent Redundant Particles
        \item Reduces Quantization Error
    \end{itemize}
\end{itemize}
\end{column}

\begin{column}{.5\textwidth}
\includegraphics[clip=true,trim=0cm 6cm 0cm 6cm,width=\textwidth]{particle_filter2}
\end{column}
\end{columns}
\end{frame}

\begin{frame}{Construction}
\begin{itemize}
    \item Mixture PDF at time $k$, with $N_p$ particles:\\
        $$P(x_k) = \sum_{i=0}^{N_p} w^i\delta(x_k - x^i_k ) dx$$
    \item Weight of Particle $i$, at time $t_k$:\\
        $$w^i_k \propto \frac{P(x^i_{0:k} | y_{0:k})}{q(x^i_{0:k} | y_{0:k})}$$
      \note{$q$ is the density/location of the particle}
      \note{$p$ is the actual probability }
    \item Incorporating a Measurement, $y_k$:\\
        $$w^i_k \propto & w^i_{k-1}P(y_k| x_k) $$
      \note{Because of the proportion, note that the weights do not
            have to integrate to 1, and can be extremely large or extremely small}
\end{itemize}
\end{frame}

\section{Methods}
\begin{frame}{Particle Filter Setup}
\footnotesize
\begin{itemize}
    \item Weighting function
        \begin{itemize}
            \item Continuous, Long Tailed, Zero-Mean 
            \item Too wide $\rightarrow$ under-sensitivity, 
                slow or no convergence
            \item Too thin $\rightarrow$ reduces robustness 
                to noise, particle deprivation
            \item For this work, $N(0, 0.005^2)$ used
        \end{itemize}
    \item Resampling 
        \begin{itemize}
            \item Stratified Resampling can result in truncated tails 
                    on posterior
            \item Regularized Resampling can result in over smoothing and 
                    slow convergence
            \item Resampling Performed Rarely
        \end{itemize}
    \item Number of particles
        \begin{itemize}
            \item More particles give higher fidelity 
            \item Large Initial Particle Count (16000)
        \end{itemize}
    \item Prior Distribution
    \begin{itemize}
            \item Large $\tau$ parameters can oversmooth
            \item $\epsilon$ needs large variance
    \end{itemize}
\end{itemize}
\note{Because the first 20 measurements drop most weights very quickly this gives a large breadth,
    but after the initial calculations drops the computation requirements}
\note{This has not been done before}
\end{frame}

\begin{frame}{FMRI Noise}
\centering

@@

\section{Conclusion}
\begin{frame}{Conclusion}
\begin{itemize}
    \item Summary:
    \begin{itemize}
    \item BOLD Parameters Ill-Defined
    \item Particle Filter Capable of good parameter BOLD estimate with 1000 particles
    \item Mutual Information performs well as estimate of Quality
    \end{itemize}
    \item Future Works
    \begin{itemize}
        \item Further limitations should be placed on priors, Deneux 
            et al. \cite{Deneux2006} shows that parameters could imposed.
        \item Analysis of joint parameter distribution for populations
        \item Real Time analysis possible for multiple Voxels, similar to
            De Charms et al. \cite{DeCharms2005}
